Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-11 paper/main.tex
some rephrasing and starting on Augustus stuff

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we do want to train with as much data as possible, but 
not all data has a good enough quality to enable a powerful generalization. It is 
also very desirable to have a broad spread phylogenetically as well as in genome 
size and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, if the genomic patterns 
inside the data are too different from each other, it may be difficult for the model 
to generalize, too. It was also important for us to get experimental results within 
a couple of days as we tested many different data and model configurations, which 
limited the overall size of the training set and the model.

@@

Once a set of training genomes was selected, we split it up again into a {\it 
training and validation set}. This split of the training genomes was done to get a 
quicker sense of the generalization capabilities and was used primarily during 
training. We split off the validation set by selecting 20\% of the FASTA sequences 
above and below the N90 of each training genome, ensuring a proper distribution of 
large and short sequences in both sets and a split on the chromosome level. 
Evaluation of the annotations on all training and validation genomes was done 
regularly after a promising model candidate was found based on its validation set 
performance. Ultimately we only cared about the cross-species performance on all 
validation genomes of a model.

@@

\begin{table}[!t]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{{\it Note:} Values are averages of the individual values of each 
genome in a data group. All statistics except the average gene lengths are excluding 
FASTA sequences without a gene and subsequences that were masked as completely 
erroneous. Each strand of DNA was counted seperately. The average gene length was 
determined by the average length of the longest transcript by cumulative exon 
length, which is how we query for genes during data generation. Brackets show the 
standart deviation.}
\end{table}

During data generation, we query for the longest transcript of each gene and 
disregard FASTA sequences that have no functional annotation. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of bases masked lie in the intergenic region as the most prevelant error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continous genomic sequence into 20.000 bases 
long subsequences, which are then used as input or label for the neural network 
together with the sample weights. We append zero padding if the subsequences are 
not long enough. If a subsequence is fully marked as erroneous, we do not include 
it anywhere.

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
and further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Both metrics do not take the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S1 and S2 for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the classes of the input data are not balanced equally enough, at 
least not for our network, and does also vary greatly between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. A class wise evaluation including the calculation of the Genic 
F1 score is performed on the validation set after every epoch.

We also implemented a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section 6 of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
larger, depending on species group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The sequence length during inference can 
be up to 200.000 base pairs long and is determined by the phylogenetic class. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section 3 of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cut to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were recombined by averaging 
the individual softmax values of each base. The figures in Section 4 of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we trained 4 seperate animal and plant models and also used 
a model from an epoch that did not have the highest Genic F1 on the validation set. 
Depending on the results, we used the model iteration with either the highest genic 
recall or precision. The model with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
What Marvin did, why we don't use Genic F1 for comparison, etc

>>>>

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we do want to train with as much data as possible, but 
not all data has a good enough quality to enable a powerful generalization. It is 
also very desirable to have a broad spread phylogenetically as well as in genome 
size and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize, if the genomic patterns inside the data are 
too different from each other. It was also important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

Once a set of training genomes was selected, we further split the sequences therein into a {\it 
training and validation set}. This split of the training genomes was done to get a 
quicker sense of the generalization capabilities and was used primarily during 
training. We split off the validation set by selecting 20\% of the FASTA sequences 
above and below the N90 of each training genome, ensuring a proper distribution of 
large and short sequences in both sets and a split on the chromosome level. 
Evaluation of the annotations on all training and validation genomes was done 
regularly after a promising model candidate was found based on its validation set 
performance. Ultimately we used the cross-species performance on all 
validation genomes as the decisive measure of model quality.

@@

\begin{table}[!t]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{{\it Note:} Values are averages of the individual values of each 
genome in a data group. All statistics except the average gene lengths exclude
FASTA sequences without a gene and any 20,000 bp subsequences that were masked as completely 
erroneous. Each strand of DNA was counted seperately. The gene length was 
determined by the length of the pre-mRNA of the longest protein at each
loci. Brackets show the standard deviation.}
\end{table}

During data generation, we query for the transcript with the longest protein of each gene and 
disregard FASTA sequences that have no functional annotation as it is ambiguous
whether such sequences contain no genes or were simply not annotated in the reference. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of masked bases lie in the intergenic region as the most prevelant error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the sequences (or
sequence ends) are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
and further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S1 and S2 for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes. %?? hmmm, but this is not a weighted mean across classes is it? Do introns not weigh more heavily?

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the classes of the input data are both unbalanced vary greatly 
between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

A class wise evaluation including the calculation of the Genic % still could use a better location...
F1 score is performed on the validation set after every epoch. The best model
from any training run was selected as the model with the highest Genic F1 after either
maximum epochs were reached or Genic F1 stopped improving and training was interrupted. % check / what was patience set to?

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section 6 of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section 3 of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section 4 of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two models each, specifically the model with the best precision and the  % please check
other with the best recall. The model with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, Augustus.
Typical usage requires retraining Augustus for each species, with the exception of a
few lucky cases where a model for a close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

Orthologs of highly conserved single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants.

What Marvin did, why we don't use Genic F1 for comparison, etc
